---
title: "A2 Project 2"
author: "Johnsen & Johnsen"
date: "`r Sys.Date()`"
output: pdf_document
---
```{r, include = F}
knitr::opts_chunk$set(warning = F, fig.height = 4, message = F, dpi = 500)
rm(list = ls())
library(lubridate)
library(latex2exp)
library(circular)
library(tidyverse)
library(reshape)
library(pheatmap)
library(openair)
library(stringr)
library(numDeriv)
library(gridExtra)
library(openair)
library(PowerNormal)
library(sn)
library(gnorm)
library(emg)
library(survival)
library(survminer)


if (Sys.getenv("LOGNAME") == "mortenjohnsen"){
  setwd("/Users/mortenjohnsen/OneDrive - Danmarks Tekniske Universitet/DTU/9. Semester/02418 - Statistical Modelling/Project-1/")
} else {
  setwd("~/Documents/02418 Statistical Modelling/Assignments/Assignment 1/Project-1")
}

source("testDistribution.R")
```

# Projekt 2: Survival Data

## Analysis of the Binary Data

### Read the data Logistic.txt into R.

```{r}
log.data <- read.table("Logistic.txt", header=TRUE, sep="", 
                       as.is=TRUE)
```


### Fit a logistic regression model for the binary outcome AIDS=”yes” versus AIDS=”no” with the explanatory variable treatment with AZT (Yes, NO). Present the odds ratio for the effect of AZT on AIDS with 95% confidence interval and interpret the result in words

### Test the hypothesis of no effect of AZT on AIDS using:

* The likelihood ratio test
* The Wald test
* The score test


## Analysis of the survival time data

### Descriptive statistics

#### Read the data actg320.txt into R

#### How many patients got AIDS or died in the two treatment groups? And how long was the total follow-up time in the two groups?

#### Plot the survival functions in the two treatment groups, which group seems to be doing best?

#### Plot the cumulative incidence functions for the two groups, which plot would you prefer?

#### Compare the survival in the two treatment groups using a log-rank test.


### Parametric survival models

#### Fit parametric survival models containing treatment (tx) and CD4 count (cd4) as explanatory variables

* Try using the exponential, Weibull and log-logistic models, which one gave the best fit (and why)?   


#### Using the survival model you chose, make a table of estimates and their 95% confidence intervals


#### Using your model compute the time ratio for the treatment effect. Sim- ilarly, compute the time ratio for the effect of increasing the CD4 count with 50. In both cases unceartainty evaluation (e.g. confidence inter- vals) should be included. Interpret the results in words


#### Assess the goodness of fit of this model using a plot based on the Cox Snell residuals


#### Give a graphical presentation of your model