---
title: "A2 Project 1"
author: "Johnsen & Johnsen"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

Setting working directories and loading the data. Data is also normalized.

```{r}
library(ggplot2)
library(numDeriv)
library(betareg)
library(gridExtra)

if (Sys.getenv("LOGNAME") == "mortenjohnsen"){
  setwd("/Users/mortenjohnsen/OneDrive - Danmarks Tekniske Universitet/DTU/9. Semester/02418 - Statistical Modelling/Project-1/")
} else {
  setwd("~/Documents/02418 Statistical Modelling/Assignments/Assignment 1/Project-1")
  #setwd("~/Documents/02418 Statistical Modelling/Assignments/Assignment 1/Project-1/Project 2")
  #source("testDistribution.R")
}

D <- read.table("tuno.txt", header=TRUE, sep=" ", 
                as.is=TRUE)

D$date <- as.Date("2003-01-01")-1+D$r.day
D$pow.obs.norm <- D$pow.obs/5000
```

Data is transformed to be normal using transformation 1 in description of assignment 1:
$$
y^{(\lambda)}=\frac{1}{\lambda}log\left(\frac{y^\lambda}{1-y^\lambda}\right); \lambda > 0
$$
Below can be seen qq-plots comparing the normalized data before and after transformation to allow for a better understanding of the motivation behind the transformation.

```{r}
#lambdas <- seq(-0.5,0.5,by=0.01)
#library(MASS)
#boxcox(lm(D$pow.obs.norm~1), lambda=lambdas)

#Define transformation function
Trans.eq1 <- function(lambda, y){
  y_lambda <- 1/lambda * log(y^lambda/(1-y^lambda))#, lambda > 0
  return(y_lambda)
}
#Optimization function
#Måske er det bedre at lave nogle undersøgelser selv frem for bare at optimere lambda (Overvej til senere).
#se kode fra lecture 4 linje 5-73
lambda_NLL <- function(lambda, x = D$pow.obs.norm){
  y <- Trans.eq1(lambda, x)
  NLL <- -as.numeric(shapiro.test(y)$statistic)
  return(NLL)
}
lambda.hat <- nlminb(start = 0.2, objective = lambda_NLL)
#round to two decimal points.
lambda <- round(lambda.hat$par, 2)
D$transformed.pow.obs.norm <- Trans.eq1(lambda, D$pow.obs.norm)

#Check qqplot:
par(mfrow = c(2,2))
qqnorm(D$pow.obs.norm)
qqline(D$pow.obs.norm)
hist(D$pow.obs.norm)
qqnorm(D$transformed.pow.obs.norm)
qqline(D$transformed.pow.obs.norm)
hist(D$transformed.pow.obs.norm)
```
Two models will be produced: a non-normal beta or gamma regression model or the normalized data and a normal linear regression model of the transformed, normalized data.

First the non-normal:
```{r}
#gamma's pretty bad. Maybe beta would be better?
gammaModel <- glm( (pow.obs.norm ~ ws30 + I(ws30^2)), data = D, family = Gamma)
betaModel <- betareg(formula = (pow.obs.norm ~ ws30 + I(ws30^2)), data = D)
D$gammaModel.pow <- gammaModel$fitted.values
D$betaModel.pow <- betaModel$fitted.values

ggplot(data = D)+
  geom_point(aes(x=ws30, y=pow.obs.norm, colour="Data"))+
  geom_line(aes(x=ws30, y=gammaModel.pow, colour="Gamma"))+
  geom_line(aes(x=ws30, y=betaModel.pow, colour="Beta"))+
  labs(x = "Wind speeds", y="Norm power obs", colour = "")+
  theme_bw()+
  ggtitle("Models in standardised domain")
```
This can also be done by hand. See p. 163-65 for reference. Exponential family regression models estimate the log-likelihood contribution of an outcome $y_i$ as seen below:
$$
log L(\theta_i,\phi)=\frac{y_i\theta_i-A(\theta_i)}{\phi}+c(y_i,\phi)
$$
Here, $A(\theta_i)$ is defined as such that its derivative is the mean $E_{y_1}=A'(\theta_i)=\mu_i$. $\phi$ is the dispersion parameter, where $var(y_i)=\phi A''(\theta_i)=\phi v(\mu_i)$. To use this general exponential family for regression analysis we only have to specify $A(\theta_i)$ and a link function $h(\mu_i)$. The reason as to why $c(y_i,\phi)$ can be neglected is that the term disappears when determining the score function anyways.??? KOMMENTAR


Now to the normal linear model:

```{r cars}
ws.only <- lm( (D$transformed.pow.obs.norm ~ ws30), data = D ) #good for comparison transfit1
ws.and.ws.squared <- lm( (D$transformed.pow.obs.norm ~ ws30 + I(ws30^2)), data = D) #brilliant transfit12
ws.squared.only <- lm( (D$transformed.pow.obs.norm ~ I(ws30^2)), data = D ) #1. jeg har ladt disse to stå for at illustrere transfit2 noget med residualerne
ws.ws2.and.wd <- lm( (D$transformed.pow.obs.norm ~ ws30 + I(ws30^3) + cos(wd30)), data = D ) #2. 
summary(ws.ws2.and.wd)
#transfitS <- lm( (D$transformed.pow.obs.norm ~ ws30 + I(ws30^2) + Season), data = D )
#summary(ws.only)
#summary(ws.and.ws.squared)

D$ws.pred <- ws.only$coefficients[1] + ws.only$coefficients[2] * D$ws30
D$ws.ws2.pred <- ws.and.ws.squared$coefficients[1] + ws.and.ws.squared$coefficients[2] * D$ws30 + ws.and.ws.squared$coefficients[3] * D$ws30^2
D$ws.ws2.wd.pred <- ws.ws2.and.wd$coefficients[1] + ws.ws2.and.wd$coefficients[2] * D$ws30 + 
  ws.ws2.and.wd$coefficients[3] * D$ws30^2 + ws.ws2.and.wd$coefficients[4] * cos(D$wd30)

#D$transformed.predS <- transfitS$coefficients[1] + transfitS$coefficients[2] * D$ws30 + transfitS$coefficients[3] * D$ws30^2 + transfitS$coefficients[4] * D$Season
#summary(transfitS)

ggplot(data = D)+
  geom_point(aes(x=ws30, y=transformed.pow.obs.norm, colour="Data"), size = 4, alpha = 0.6)+
  geom_line(aes(x=ws30, y=ws.pred, colour="ws only"))+
  geom_line(aes(x=ws30, y=ws.ws2.pred, colour="ws + ws^2"))+
  scale_colour_manual(values = c("red", "blue", "orange"))+
  #geom_line(aes(x=ws30, y=ws.ws2.wd.pred, colour="ws + ws^2 + cos(wd)"))+
  labs(x = "Wind speeds", y="Transformed, norm power obs", colour = "")+
  theme_bw()+
  ggtitle("Fitted models in Transformed Domain")

summary(lm( (D$transformed.pow.obs.norm ~ ws30 + I(ws30^2) + wd30), data = D)) #simply a check with the final model to make sure that wd still is irrelevant as regressor
```
```{r}
#Outcommented so it is possible to create a PDF
# library(plotly)
# 
# fig <- plot_ly(mtcars, x = D$ws30, y = cos(D$wd30), z = D$transformed.pow.obs.norm
#                , color = "Observed Power Production (Normalised)", colors = "red")
# fig <- fig %>% add_markers()
# fig <- fig %>% layout(scene = list(xaxis = list(title = 'Pos.Obs'),
#                                    yaxis = list(title = 'cos(WD)'),
#                                    zaxis = list(title = 'WS')))
# add_trace(type = 'mesh3d',
#           p = fig,
#           x = D$ws30,
#           y = cos(D$wd30),
#           z = D$ws.ws2.wd.pred,
#           facecolor = rep("blue", length(D$ws30)),
#           opacity = 0.1
# )
```

Tror ikke at det giver mening at have wind direction med, det ligner ikke rigtig at det giver så meget? Men AIC er dog bedre... Ligeledes er det også statistisk signifikant, hvis man tilføjer måned på året.  

Tjek af modellerne:

```{r pressure, echo=FALSE, fig.height = 6, fig.width = 6}
par(mfrow=c(2,2))
plot(D$ws30, ws.only$residuals, main = "ws only")
plot(D$ws30, ws.squared.only$residuals, main = "ws^2 only")
plot(D$ws30, ws.ws2.and.wd$residuals, main = "ws, ws^2 and wd")
plot(D$ws30, ws.and.ws.squared$residuals, main = "ws + ws^2")
```

Normal linear model by hand and by optimisation:
```{r}
D$base <- 1 #slide 15 for all of this
X <- as.matrix(data.frame(D$base, D$ws30, D$ws30^2))
Y <- D$transformed.pow.obs.norm
betas <- solve( (t(X)%*%X) ) %*% t(X) %*% Y

NLL.norm.lin <- function(p, regr1, regr2, response){
  mu <- p[1] + p[2]*regr1 + p[3]*regr2
  sigma <- sqrt( 1/length(response) * sum( (response - mu)^2  ) )
  return( -sum( dnorm( x = response, mean = mu, sd = sigma, log = T ) )  )
}
norm.lin.par <- nlminb(c(1,1,1), objective = NLL.norm.lin, regr1 = D$ws30, regr2 = D$ws30^2, response = Y)

round( rbind(ws.and.ws.squared$coefficients,betas[1:3],norm.lin.par$par), digits=6)
```
```{r, fig.width=9, fig.height=3}
y_p <- D$ws.ws2.pred
#y_pS <- D$transformed.predS
D$y_inv_trans <- 1/( exp(y_p*lambda)+1 )^(1/lambda) * exp(y_p)
#D$Season_inv_trans <- 1/( exp(y_pS*lambda)+1 )^(1/lambda) * exp(y_pS)

#D$Season <- 0
#D$Season[D$month > 5] <- 1

fitplot <- ggplot(data = D)+
  geom_point(aes(x=ws30, y=pow.obs.norm, colour = "Observed power production (Transformed)"), size = 4, alpha = 0.6)+
  geom_line(aes(x=ws30, y=y_inv_trans, colour="Inverse Transformation of the Normal model"), size = 2)+
  geom_line(aes(x=ws30, y=betaModel.pow, colour="The directly fitted beta model"), size = 2)+
  #geom_line(aes(x=ws30, y=Season_inv_trans, colour="InvS"))+
  labs(x = "Wind speeds", y="Norm power obs", colour = "")+
  theme_bw()+
  labs(colour = "")+
  ggtitle("Model fits")+
  theme(legend.background = element_rect(colour = "black"), legend.position = "top")+
  scale_colour_manual(values = c("orange", "red", "blue"))
#scale_shape_manual(values = c(1:10))+
#scale_colour_manual(values = c("blue", "yellow", "black"))

resplot <- ggplot(data = D)+
  geom_point(aes(x = ws30, y = pow.obs.norm - y_inv_trans, colour = "Normal model residuals"), size = 3, alpha = 0.5)+
  geom_point(aes(x = ws30, y = pow.obs.norm - betaModel.pow, colour = "Beta model residuals"), size = 3, alpha = 0.5)+
  theme_bw()+
  scale_colour_manual(values = c("blue", "orange"))+
  ggtitle("Model Residuals Comparison")+
  labs(colour = "")+
  theme(legend.position = "top", legend.background = element_rect(colour = "black"))

grid.arrange(fitplot,resplot, nrow = 1)

#plot(D$ws30, D$pow.obs.norm-D$y_inv_trans)
#abline(h=0)

```

```{r}
AIC(betaModel)
AIC(ws.and.ws.squared)
logLik(betaModel)
g_inverse <- function(y,lambda){
  return((1/(exp(y*lambda)+1))^(1/lambda) * exp(y))
}
#2*norm.lin.par$objective+2*4
norm.lin.par$objective-sum(log(-1/(D$pow.obs.norm*(-1+D$pow.obs.norm^0.26))))
#AIC(ws.and.ws.squared)-sum(log(-1/(D$pow.obs.norm*(-1*D$pow.obs.norm^0.26))))
```


Uncertainties of parameters (in transformed domain)
```{r}
norm.lin.sds <- sqrt(diag(solve(hessian( func = NLL.norm.lin, x = norm.lin.par$par, regr1 = D$ws30, regr2 = D$ws30^2, response = D$transformed.pow.obs.norm ))))
Wald.CI.upper <- norm.lin.par$par + qnorm(0.975) * norm.lin.sds
Wald.CI.lower <- norm.lin.par$par - qnorm(0.975) * norm.lin.sds

cat("beta_0 = ", norm.lin.par$par[1], "95% CI [",Wald.CI.lower[1],", ",Wald.CI.upper[1],"]"
    ,"\nbeta_1 = ", norm.lin.par$par[2], "95% CI [",Wald.CI.lower[2],", ",Wald.CI.upper[2],"]"
    ,"\nbeta_2 = ", norm.lin.par$par[3], "95% CI [",Wald.CI.lower[3],", ",Wald.CI.upper[3],"]"
)

#Test regularity
score.nll.norm <- function(theta, regr1, regr2, response, sigma){
  
  beta0 <- theta[1]
  beta1 <- theta[2]
  beta2 <- theta[3]
  n <- length(response)
  
  theta <- beta0 + beta1 * regr1 + beta2 * regr2
  #sigma <- sqrt( 1/length(response) * sum( (response - theta)^2 ) )
  score <- n/sigma^2 * (mean(response) - theta)#p. 213
  return(-sum(score))
}

beta0.test.range <- seq(Wald.CI.lower[1], Wald.CI.upper[1], abs(0.01*norm.lin.par$par[1]))
beta1.test.range <- seq(Wald.CI.lower[2], Wald.CI.upper[2], abs(0.01*norm.lin.par$par[2]))
beta2.test.range <- seq(Wald.CI.lower[3], Wald.CI.upper[3], abs(0.01*norm.lin.par$par[3]))

sigma.s <- sqrt( 1/dim(D)[1] * sum ( (D$transformed.pow.obs.norm - ( norm.lin.par$par[1] + norm.lin.par$par[2] * D$ws30 + norm.lin.par$par[3] * D$ws30^2) )^2 ) )

res0 <- apply(cbind(beta0.test.range, norm.lin.par$par[2], norm.lin.par$par[3]), MARGIN = 1, FUN = score.nll.norm, regr1 = D$ws30, regr2 = D$ws30^2, response = D$transformed.pow.obs.norm, sigma = sigma.s)

res1 <- apply(cbind(norm.lin.par$par[1], beta1.test.range, norm.lin.par$par[3]), MARGIN = 1, FUN = score.nll.norm, regr1 = D$ws30, regr2 = D$ws30^2, response = D$transformed.pow.obs.norm, sigma = sigma.s)

res2 <- apply(cbind(norm.lin.par$par[1], norm.lin.par$par[2], beta2.test.range), MARGIN = 1, FUN = score.nll.norm, regr1 = D$ws30, regr2 = D$ws30^2, response = D$transformed.pow.obs.norm, sigma = sigma.s)

par(mfrow = c(1,3))
plot(beta0.test.range, res0, main = "Score linearity for beta0"
     ,xlab = "beta0", ylab = "Score", pch = 16)
abline(lm(res0 ~ beta0.test.range), col = 2, lwd = 2)
grid()

plot(beta1.test.range, res1, main = "Score linearity for beta1"
     ,xlab = "beta1", ylab = "Score", pch = 16)
abline(lm(res1 ~ beta1.test.range), col = 2, lwd = 2)
grid()

plot(beta2.test.range, res2, main = "Score linearity for beta2"
     ,xlab = "beta2", ylab = "Score", pch = 16)
abline(lm(res2 ~ beta2.test.range), col = 2, lwd = 2)
grid()

```

Interpretation of parameters 
```{r}
ws.and.ws.squared$coefficients
beta_0 <- ws.and.ws.squared$coefficients[1]
( offset_inv <- 1/( exp(beta_0*lambda)+1 )^(1/lambda) * exp(beta_0) )
```
The regression model used is a normal model with data transformation. The first parameter is the intercept with the y-axis and can be transformed back to the original domain yielding an offset value of 0.00035. This is as expected since there should be no power production at 0 wind speed.
The other two parameters are slope coefficients. The first is positive and describes the relationship between the wind speed and the power production. According to this model it should increase linearly (in the transformed domain). The second coefficient is negative and implies a negative proportionality to the square of the wind speed (in the transformed domain).

