---
title: "A1 Project 3"
author: "Johnsen & Johnsen"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r, include = F}
knitr::opts_chunk$set(warning = F, fig.height = 4, message = F, dpi = 500)
rm(list = ls())
library(lubridate)
library(latex2exp)
library(circular)
library(tidyverse)
library(reshape)
library(pheatmap)
library(openair)
library(stringr)
library(numDeriv)
library(gridExtra)
library(openair)
library(PowerNormal)
library(sn)
library(gnorm)
library(emg)
library(survival)
library(survminer)


if (Sys.getenv("LOGNAME") == "mortenjohnsen"){
  setwd("/Users/mortenjohnsen/OneDrive - Danmarks Tekniske Universitet/DTU/9. Semester/02418 - Statistical Modelling/Project-1/")
} else {
  setwd("~/Documents/02418 Statistical Modelling/Assignments/Assignment 1/Project-1")
}
```

## Projekt 3: Financial Data

### Descriptive Statistics and Simple Models

#### Present the data, estimate the parameters in a normal model, and asses if the normal model is appropriate.

```{r}
D <- read.table("finance_data.csv", header=TRUE, sep=";", 
                as.is=TRUE)
## Dimensions of D (number of rows and columns)
dim(D)
##  Column/variable names
names(D)
## The first rows/observations
head(D)
## The last rows/observations
tail(D)
## Selected summary statistics
summary(D)
## Another type of summary of the dataset
str(D)

D$date <- as.Date(D$time)
D$year <- year(D$date)

ggplot(D, aes(x = date, y = SLV)) + geom_point()

# ggplot(D, aes(x = SLV)) +
#   geom_histogram(aes(y = ..density..), color = 'black')

par <- nlminb(start = c(1,1), objective = testDistribution,
              distribution = "normal",
              x = D$SLV)
ggplot(D)+
  geom_histogram(aes(x = SLV, y= ..density..,), color='black') +#color, fill
  stat_function(fun = dnorm, n = dim(D)[1], args = list(mean = par$par[1], sd = par$par[2]), color='red')

# plot(ecdf(D$SLV), verticals = T)
# xseq <- seq(0.9*min(D$SLV), 1.1*max(D$SLV), length.out=100)
# lines(xseq, pnorm(xseq, mean(D$SLV), sd(D$SLV)), col='red')
#plot(xseq, pnorm(xseq, mean(D$SLV), sd(D$SLV)), col='red')
qqnorm(D$SLV)
qqline(D$SLV)
```


#### Hypothesize a model that could fit the data better (Hint: consider tail probabilities), and compare with the normal model estimated above

```{r}
lcauchyFUNC <- function(p, data){
  x0 <- p[1] #location R
  gam <- p[2] #scale R > 0
  return(-sum(dcauchy(x = data, location = x0, scale = gam, log = T)))
}
lpownormFUNC <- function(p, data){
  alpha <- p
  return(-sum(log(dpn(x = data, p))))
}
ltFUNC <- function(p, data){
  return(-sum(dt(x = data, df = p, log = T)))
}
lsnFUNC <- function(p, data){ #skewed normal dist
  return(-sum(dsn(x = data, xi = p[1], omega = p[2], alpha = p[3], log = T)))
}
lgnFUNC <- function(p, data){ #symmetric generalized normal dist
  return(-sum(dgnorm(x = data, mu = p[1], alpha = p[2], beta = p[3], log = T)))
}

lasgnFUNC <- function(p, data){ #asymmetric generalized normal dist, when K = 0 has already been checked
  epsilon <- p[1]
  alpha <- p[2]
  kappa <- p[3]
  return(-sum( log(dnorm(x = -1/kappa * log(1 - kappa * (data - epsilon) / alpha) ) /
              (alpha - kappa * (data - epsilon)) ) ) )
}

lemgFUNC <- function(p, data){ #exponential modified gaussian dist
  return(-sum(demg(x = data, mu = p[1], sigma = p[2], lambda = p[3], log = T)))
}

par.cauchy <- nlminb(start = c(0,1), objective = lcauchyFUNC, data = D$SLV)
par.pownorm <- nlminb(start = 1, objective = lpownormFUNC, data = D$SLV)
par.t <- nlminb(start = 1, objective = ltFUNC, data = D$SLV)
par.sn <- nlminb(start = c(1,1,1), objective = lsnFUNC, data = D$SLV)
par.gn <- nlminb(start = c(1,1,1), objective = lgnFUNC, data = D$SLV)
par.asgn <- nlminb(start = c(1,1,1), lower = c(-Inf, -Inf, 0), objective = lasgnFUNC, data = D$SLV)
par.emg <- nlminb(start = c(1,1,1), lower = c(-Inf, 1/1000, 1/1000), objective = lemgFUNC, data = D$SLV)

ggplot(D)+
  geom_histogram(aes(x = SLV, y= ..density..,), color='black') + #color, fill
  stat_function(fun = dnorm, n = dim(D)[1], args = list(mean = par$par[1], sd = par$par[2]), aes(colour = "norm")) +
  stat_function(fun = dcauchy, n = dim(D)[1], args = list(location = par.cauchy$par[1],
                                                          scale = par.cauchy$par[2]), aes(colour = "cauchy")) +
  stat_function(fun = dpn, n = dim(D)[1], args = list(alpha = par.pownorm$par), aes(colour = "powernorm")) +
  stat_function(fun = dt, n = dim(D)[1], args = list(df = par.t$par), aes(colour = "t")) +
  stat_function(fun = dsn, n = dim(D)[1], args = list(xi = par.sn$par[1], omega = par.sn$par[2],
                                                      alpha = par.sn$par[3]), aes(colour = "sn")) +
  stat_function(fun = dgnorm, n = dim(D)[1], args = list(mu = par.gn$par[1], alpha = par.gn$par[2],
                                                         beta = par.gn$par[3]), aes(colour = "gnorm")) +
  stat_function(fun = demg, n = dim(D)[1], args = list(mu = par.emg$par[1], sigma = par.emg$par[2],
                                                       lambda = par.emg$par[3]), aes(colour = "emg"))+
  scale_colour_manual(values = c("blue", "red", "yellow", "black", "pink", "grey", "purple"))+
  labs(colour = "Distribution")
#legend('topright', legend=c('normal', 'cauchy', 'power normal', 't'), col=c('red', 'blue', 'green', 'yellow'))

AIC.norm <- -2 * sum(dnorm(x = D$SLV, mean = par$par[1], sd = par$par[2], log = T))
+ 2 * length(par$par)
AIC.cauchy <- -2 * sum(dcauchy(x = D$SLV, location = par.cauchy$par[1],
                               scale = par.cauchy$par[2], log = T)) + 2 * length(par.cauchy$par)
AIC.pownorm <- -2 * sum(log(dpn(x = D$SLV, alpha = par.pownorm$par)))
+ 2 * length(par.pownorm$par)
AIC.t <- -2 * sum(dt(x=D$SLV, df = par.t$par, log = T)) + 2 * length(par.t$par)
AIC.sn <- -2 * sum(dsn(x=D$SLV, xi = par.sn$par[1], omega = par.sn$par[2], alpha = par.sn$par[3], 
                       log = T)) + 2 * length(par.sn$par)
AIC.gn <- -2 * sum(dgnorm(x=D$SLV, mu = par.gn$par[1], alpha = par.gn$par[2], beta = par.gn$par[3], 
                          log = T)) + 2 * length(par.gn$par)
AIC.asgn <- -2 * sum( log(dnorm(x = -1/par.asgn$par[3] * log(1 - par.asgn$par[3] * (D$SLV - par.asgn$par[1]) / par.asgn$par[2]) ) /
                            (par.asgn$par[2] - par.asgn$par[3] * (D$SLV - par.asgn$par[1])) ) ) + 2 * length(par.asgn$par)
AIC.emg <- -2 * sum(demg(x=D$SLV, mu = par.emg$par[1], sigma = par.emg$par[2], lambda = par.emg$par[3], 
                         log = T)) + 2 * length(par.emg$par)


round(rbind(AIC.norm, AIC.cauchy, AIC.pownorm, AIC.t, AIC.sn, AIC.gn, AIC.asgn, AIC.emg), digits=5)

n <- 100000 
par(mfrow=c(2,3)) #comparing by means of generating histograms with n = 100000 points for each of the distributions
hist(rnorm(n, mean = par$par[1], sd = par$par[2]))
hist(D$SLV)
hist(rsn(n, xi = par.sn$par[1], omega = par.sn$par[2], alpha = par.sn$par[3]))
hist(rgnorm(n, mu = par.gn$par[1], alpha = par.gn$par[2], beta = par.gn$par[3]))
hist(rcauchy(n, location = par.cauchy$par[1], scale = par.cauchy$par[2]))
hist(remg(n, mu = par.emg$par[1], sigma = par.emg$par[2], lambda = par.emg$par[3]))

par(mfrow=c(1,3)) #comparing by means of their empirical distribution functions
plot(ecdf(D$SLV), verticals = T) 
xseq <- seq(0.9*min(D$SLV), 1.1*max(D$SLV), length.out=100)
#lines(xseq, pnorm(xseq, mean(D$SLV), sd(D$SLV)), col='red')
lines(xseq, pnorm(xseq, mean = par$par[1], sd = par$par[2]), col='blue')
plot(ecdf(D$SLV), verticals = T)
lines(xseq, pgnorm(xseq, mu = par.gn$par[1], alpha = par.gn$par[2], beta = par.gn$par[3]), col='green')
plot(ecdf(D$SLV), verticals = T)
lines(xseq, psn(xseq, xi = par.sn$par[1], omega = par.sn$par[2], alpha = par.sn$par[3]), col='red')

```


#### Present the final model (i.e. relevant keynumbers for the estimates)
First we have the profile likelihoods of the two parameters of the normal distribution optimized at the start of project 3. This is to enable comparison with the profile likelihoods of the chosen distribution which is the generalized normal distribution. Likelihood based and Wald CIs are also calculated but are not printed till later.

```{r}
alpha <- 0.05
c <- exp(-0.5 * qchisq(1-alpha, df = 1))
par(mfrow=c(1,2))
#likelihood-based CI for normal distribution
mle.norm <- par$par

fun.norm <- function(mu, sigma, data){#####
  prod(dnorm(x = data, mean = mu, sd = sigma, log = F) / 2) #to avoid inf-values
}

l.fun.norm <- function(mu, sigma, data){#####
  sum(dnorm(x = data, mean = mu, sd = sigma, log = T))
}

CIfun.norm <- function(y, data, mu = T){##### T from mean, F for sigma
  if(mu){
    sum(dnorm(x = data, mean = mle.norm[1], sd = mle.norm[2], log = T)) -
      sum(dnorm(x = data, mean = y, sd = mle.norm[2], log = T)) - 
      0.5 * qchisq(1-alpha, df = 1)
  } else {
    sum(dnorm(x = data, mean = mle.norm[1], sd = mle.norm[2], log = T)) -
      sum(dnorm(x = data, mean = mle.norm[1], sd = y, log = T)) - 
      0.5 * qchisq(1-alpha, df = 1) 
  }
}
mus <- seq(-0.01, 0.013, by = 0.00001)
mu.norm <- sapply(X = mus, FUN = fun.norm, sigma = mle.norm[2], data = D$SLV)
plot(mus, mu.norm/max(mu.norm), col = 1, type = "l", xlab = expression(paste(mu)),
     main = "Parameter value for mean for normal model of SLV")
CI.mu.norm <- c(uniroot(f = CIfun.norm, interval = c(-0.003, mle.norm[1]), data = D$SLV, mu = T)$root,
                uniroot(f = CIfun.norm, interval = c(mle.norm[1], 0.006), data = D$SLV, mu = T)$root)
lines(range(mus), c*c(1,1), col = 2)

sigmas <- seq(0.033, 0.060, by = 0.00001)
sigma.norm <- sapply(X = sigmas, FUN = fun.norm, mu = mle.norm[1], data = D$SLV)
plot(sigmas^2, sigma.norm/max(sigma.norm), col = 1, type = "l", xlab = expression(paste(sigma^2)),
     main = "Parameter value for var for normal model of SLV")
CI.sigma.norm <- c(uniroot(f = CIfun.norm, interval = c(0.033, mle.norm[2]), data = D$SLV, mu = F)$root,
                   uniroot(f = CIfun.norm, interval = c(mle.norm[2], 0.063), data = D$SLV, mu = F)$root)
CI.sigmasq.norm <- CI.sigma.norm^2
lines(range(sigmas^2), c*c(1,1), col = 2)

#Wald CI for normal distribution
n <- dim(D)[1]
H.mu.norm <- hessian(l.fun.norm, mle.norm[1], sigma = mle.norm[2], data = D$SLV)
V.mu.norm <- as.numeric(-1/H.mu.norm)
H.sigma.norm <- hessian(l.fun.norm, mle.norm[2], mu = mle.norm[1], data = D$SLV)
V.sigma.norm <- as.numeric(-1/H.sigma.norm)
wald.mu.norm <- mle.norm[1] + c(-1,1) * qnorm(1-alpha/2) * sqrt(V.mu.norm)
wald.sigmasq.norm <- (mle.norm[2] + c(-1,1) * qnorm(1-alpha/2) * sqrt(V.sigma.norm))^2

mle.norm.sq <- c(mle.norm[1], mle.norm[2]^2)
```

Profile likelihoods of the three parameters of the GENERALIZED normal distribution which has been chosen as the best distribution in order to describe the ETF data. Likelihood based and Wald CIs are also calculated and printed along with those of the normal distribution.

```{r}
mle.gn <- par.gn$par
lgnFUNC <- function(p, data){ #symmetric generalized normal dist
  return(-sum(dgnorm(x = data, mu = p[1], alpha = p[2], beta = p[3], log = T)))
}

fun.Gnorm <- function(mu, alpha, beta, data){#####
  prod(dgnorm(x = data, mu = mu, alpha = alpha, beta = beta, log = F) / 2)#to avoid inf-values
}

l.fun.Gnorm <- function(mu, alpha, beta, data){#####
  sum(dgnorm(x = data, mu = mu, alpha = alpha, beta = beta, log = T))
}

CIfun.Gnorm <- function(y, data, p = "mu"){##### T from mean, F for sigma
  if(p == "mu"){
    sum(dgnorm(x = data, mu = mle.gn[1], alpha = mle.gn[2], beta = mle.gn[3], log = T)) -
      sum(dgnorm(x = data, mu = y, alpha = mle.gn[2], beta = mle.gn[3], log = T)) - 
      0.5 * qchisq(1-alpha, df = 1)
  } else if(p == "alpha") {
    sum(dgnorm(x = data, mu = mle.gn[1], alpha = mle.gn[2], beta = mle.gn[3], log = T)) -
      sum(dgnorm(x = data, mu = mle.gn[1], alpha = y, beta = mle.gn[3], log = T)) - 
      0.5 * qchisq(1-alpha, df = 1) 
  } else {
    sum(dgnorm(x = data, mu = mle.gn[1], alpha = mle.gn[2], beta = mle.gn[3], log = T)) -
      sum(dgnorm(x = data, mu = mle.gn[1], alpha = mle.gn[2], beta = y, log = T)) - 
      0.5 * qchisq(1-alpha, df = 1) 
  }
}
###PROFILE likelihoods
par(mfrow = c(1,3))
#mu
mus.gn <- seq(-0.0055, 0.011, by = 0.00001)
mu.gn <- sapply(X = mus.gn, FUN = fun.Gnorm, alpha = mle.gn[2], beta = mle.gn[3], data = D$SLV)
plot(mus.gn, mu.gn/max(mu.gn), col = 1, type = "l", xlab = expression(paste(mu)),
     main = "Parameter value for location of generalized normal model of SLV")
CI.mu.gn <- c(uniroot(f = CIfun.Gnorm, interval = c(min(mus.gn), mle.gn[1]), data = D$SLV, p = "mu")$root,
                uniroot(f = CIfun.Gnorm, interval = c(mle.gn[1], max(mus.gn)), data = D$SLV, p = "mu")$root)
lines(range(mus.gn), c*c(1,1), col = 2)
#alpha
alphas <- seq(0.042, 0.062, by = 0.0001)
alpha.gn <- sapply(X = alphas, FUN = fun.Gnorm, mu = mle.gn[1], beta = mle.gn[3], data = D$SLV)
plot(alphas, alpha.gn/max(alpha.gn), col = 1, type = "l", xlab = expression(paste(alpha)),
     main = "Parameter value for scale for generalized normal model of SLV")
CI.alpha.gn <- c(uniroot(f = CIfun.Gnorm, interval = c(min(alphas), mle.gn[2]), data = D$SLV, p = "alpha")$root,
                   uniroot(f = CIfun.Gnorm, interval = c(mle.gn[2], max(alphas)), data = D$SLV, p = "alpha")$root)
lines(range(alphas), c*c(1,1), col = 2)
#beta
betas <- seq(1.19, 1.59, by = 0.001)
beta.gn <- sapply(X = betas, FUN = fun.Gnorm, mu = mle.gn[1], alpha = mle.gn[2], data = D$SLV)
plot(betas, beta.gn/max(beta.gn), col = 1, type = "l", xlab = expression(paste(beta)),
     main = "Parameter value for shape for generalized normal model of SLV")
CI.beta.gn <- c(uniroot(f = CIfun.Gnorm, interval = c(min(betas), mle.gn[3]), data = D$SLV, p = "beta")$root,
                uniroot(f = CIfun.Gnorm, interval = c(mle.gn[3], max(betas)), data = D$SLV, p = "beta")$root)
lines(range(betas), c*c(1,1), col = 2)

#Wald CIs
n <- dim(D)[1]
H.mu.gn <- hessian(l.fun.Gnorm, mle.gn[1], alpha = mle.gn[2], beta = mle.gn[3], data = D$SLV)
V.mu.gn <- as.numeric(-1/H.mu.gn)
H.alpha.gn <- hessian(l.fun.Gnorm, mle.gn[2], mu = mle.gn[1], beta = mle.gn[3], data = D$SLV)
V.alpha.gn <- as.numeric(-1/H.alpha.gn)
H.beta.gn <- hessian(l.fun.Gnorm, mle.gn[3], mu = mle.gn[1], alpha = mle.gn[3], data = D$SLV)
V.beta.gn <- as.numeric(-1/H.beta.gn)
wald.mu.gn <- mle.gn[1] + c(-1,1) * qnorm(1-alpha/2) * sqrt(V.mu.gn)
wald.alpha.gn <- mle.gn[2] + c(-1,1) * qnorm(1-alpha/2) * sqrt(V.alpha.gn)
wald.beta.gn <- mle.gn[3] + c(-1,1) * qnorm(1-alpha/2) * sqrt(V.beta.gn)

#CIs normal model
round( rbind(CI.mu.norm, wald.mu.norm, CI.sigmasq.norm, wald.sigmasq.norm, mle.norm.sq), digits = 5)
#CIs gnorm model
round( rbind(CI.mu.gn, wald.mu.gn, CI.alpha.gn, wald.alpha.gn, CI.beta.gn, wald.beta.gn), digits=5);round(rbind (mle.gn), digits = 5)
```

Finally the expected value (and 95 % CI) of the chosen generalized normal distribution is given below.
```{r}
E.gn <- mle.gn[1]
V.gn <- mle.gn[2]^2 * gamma(3/mle.gn[3]) / gamma(1/mle.gn[3])
CI.E.gn <- E.gn + c(-1,1) * qnorm(1-alpha/2) * sqrt(V.gn / n)
rbind(E.gn);rbind(CI.E.gn)
```

The final model is compared to the distribution of the weekly returns. The parameters are shown along with the distribution model.
```{r}
temp1 <- paste("mu == ", mle.gn[1])
temp2 <- paste("alpha == ", mle.gn[2])
temp3 <- paste("beta == ", mle.gn[3])

ggplot(D)+
  geom_histogram(aes(x = SLV, y= ..density..,), color='black') + #color, fill
  stat_function(fun = dgnorm, n = dim(D)[1], args = list(mu = par.gn$par[1], alpha = par.gn$par[2],
                                                         beta = par.gn$par[3]), color = 'red') +
  annotate( "text", x = 2.7/5*max(D$SLV), y = c(10.5, 10.0, 9.4), label = c(temp1,temp2,temp3), parse = T  )+
  ggtitle("Generalized normal distribution and distribution of the weekly returns")
```

